---
title: "ST558_2024_HW8_bike data"
format: html
editor: visual
---

# Basic Modeling Practise

## Data introduction

Today we are using a bike share data for the basic modeling practise. The introoduction of this data set is accessed through this website: (Seoul Bike Sharing Demand - UCI Machine Learning Repository). To our relieve there is no missing values.

The aim of the project is to predict the rental bike count at each hour of a day based on the weather condition, season, holiday and the hour. Since bike usage is sensitive to weather condition it is presumable that certain weather condition, plus the day condition (weekday/weekend, holiday, etc.) is likely to dictate the bike usage in a particular time of the day.

Let's load some packages and read in the data first:

```{r}
library (tidyverse)
library (dplyr)
Bikeshare <- read_csv ("data//SeoulBikeData.csv")
Bikeshare

```

Unexpectedly the first trial to read in data failed. It turned out the file contains some special character that the R does not understand/not encoded by the R-recognizable way. After changing the format manually everything is fine. We first transform the date to the date format

```{r}
Bikeshare$Date <- as.Date (Bikeshare$Date, format = "%d/%m/%Y")
```

Now the date are in the right form and can do some visualization around the time data, which we will work on that later. Missing values are creating a headache for exploratory data analysis, so we want to check for missingness, evaluate and take measures to eliminate the missing values.

```{r}
colSums(is.na(Bikeshare))
length(unique(Bikeshare$Seasons))
table(Bikeshare$Seasons)

```

To check for the number of unique values in each character column, use the command table(),
